---
title: "Estuary Fish Survey - Data Rescue Mission"
subtitle: "Keystone Exercise"
author: "Ashleigh Bail"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    self-contained: true
    theme: spacelab
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r}
#| message: false
#| warning: false
#| include: false
options(repos = c(CRAN = "https://r-project.org"))

source("../code/keystone-workshop2.R")

```

### Final master dataset

Table 1 presents a preview of the final master dataset (480 observations), following standardisation of site names, taxonomic translation, removal of erroneous sensor values, integration of environmental data, and completion of zero-catch observations. The full cleaned dataset is available for download in the `outputs/tables` directory as `estuary_master_dataset.csv`.
```{r}
#| echo: true
#| message: false
#| code-fold: true

master_dataset <- complete_catch

master_dataset_display <- master_dataset |>
  mutate(
    # Make site names presentation-friendly
    site = str_replace_all(site, "_", " "),
    site = str_to_title(site),

    # Display date as day/month/year
    date = format(as.Date(date), "%d/%m/%Y"),

    # Round numeric variables for display
    across(
      c(lat, lon, mean_temperature, mean_salinity, mean_turbidity),
      ~ round(.x, 2)
    )
  ) |>
  rename(
    Site = site,
    Date = date,
    `Scientific species` = scientific_name,
    Count = count,
    Latitude = lat,
    Longitude = lon,
    Zone = zone,
    `Mean temperature (°C)` = mean_temperature,
    `Mean salinity (PSU)` = mean_salinity,
    `Mean turbidity` = mean_turbidity
  )

knitr::kable(
  head(master_dataset_display, 12),
  caption = "Table 1: Preview of the cleaned master estuary dataset.",
  align = "c"
)
```

### Statistical summary table

Table 2 presents the statistical summary of mean fish counts and salinity (± SD) for each species across estuary zones. The table highlights the variability in fish abundance and environmental conditions, providing insights into species distribution patterns within the estuarine ecosystem.
```{r}
#| echo: true
#| message: false
#| code-fold: true

catch_summary <- master_dataset |>
  group_by(scientific_name, zone) |>
  summarise(
    mean_catch = mean(count, na.rm = TRUE),
    sd_catch = sd(count, na.rm = TRUE),
    .groups = "drop"
  )

salinity_summary <- master_dataset |>
  distinct(site, date, zone, mean_salinity) |>
  group_by(zone) |>
  summarise(
    salinity_mean = mean(mean_salinity, na.rm = TRUE),
    salinity_sd = sd(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

statistical_summary_table <- catch_summary |>
  left_join(
    salinity_summary,
    by = "zone"
  )

summary_table_display <- statistical_summary_table |>
  mutate(
    zone = factor(
      zone,
      levels = c(
        "Upstream",
        "Middle",
        "Downstream",
        "Marine"
      )
    ),
    scientific_name = case_when(
      scientific_name == "Acanthopagrus spp." ~ "*Acanthopagrus* spp.",
      scientific_name == "Lates calcarifer" ~ "*Lates calcarifer*",
      scientific_name == "Lutjanus argentimaculatus" ~ "*Lutjanus argentimaculatus*",
      scientific_name == "Platycephalus spp." ~ "*Platycephalus* spp.",
      TRUE ~ scientific_name
    )
  ) |>
  arrange(scientific_name, zone) |>
  transmute(
    `Scientific species` = scientific_name,
    `Estuary zone` = zone,
    `Fish count (mean ± SD)` = sprintf(
      "%.2f ± %.2f",
      mean_catch,
      sd_catch
    ),
    `Salinity (mean ± SD)` = sprintf(
      "%.2f ± %.2f",
      salinity_mean,
      salinity_sd
    )
  )

knitr::kable(
  summary_table_display,
  align = c("l", "l", "c", "c"),
  caption = paste(
    "Table 2: Mean (± SD) fish counts and salinity",
    "for each species across estuary zones."
  )
)
```


### Multi-panel plot

```{r}
#| echo: true
#| message: false
#| code-fold: true

plot_data <- statistical_summary_table |>
  mutate(
    zone = factor(
      zone,
      levels = c(
        "Upstream",
        "Middle",
        "Downstream",
        "Marine"
      )
    )
  ) |>
  arrange(scientific_name, zone)

species_labels <- c(
  "Acanthopagrus spp." = "italic(Acanthopagrus)~spp.",
  "Lates calcarifer" = "italic(Lates~calcarifer)",
  "Lutjanus argentimaculatus" = "italic(Lutjanus~argentimaculatus)",
  "Platycephalus spp." = "italic(Platycephalus)~spp."
)

fish_plot <- ggplot(
  plot_data,
  aes(
    x = salinity_mean,
    y = mean_catch,
    group = scientific_name
  )
) +
  geom_line() +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(
      ymin = mean_catch - sd_catch,
      ymax = mean_catch + sd_catch
    ),
    width = 0.4
  ) +
  facet_wrap(
    ~ scientific_name,
    labeller = labeller(
      scientific_name = as_labeller(
        species_labels,
        label_parsed
      )
    )
  ) +
  labs(
    title = "Fish abundance along the Ross River Estuary salinity gradient" ,
    x = "Mean salinity (PSU)",
    y = "Mean fish catch (individuals per sampling day)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(hjust = 0.5)
  )

fish_plot
```

Figure 1. Mean fish abundance (± SD) along the salinity gradient of the Ross River Estuary. Panels represent individual fish taxa, with salinity increasing from upstream to marine zones.