---
title: "Workshop 4"
subtitle: "Spatial data in R"
author: "Ashleigh Bail"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    self-contained: true
    theme: spacelab
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r}
#| include: false

# Loading libraries

library(tidyverse)
library(sf)                 # simple features
library(terra)              # for raster
library(leaflet)            # interactive maps
library(tmap)               # thematic maps
```

```{r}
#| message: false
#| code-fold: true
#| echo: true
#| output: false

# Load the copepod data
dat <- read_csv("../data/workshop4/copepods_raw.csv")
dat
```

### Spatial Data Visualisation

```{r}
#| message: false
#| warning: false
#| output: true
#| code-fold: true
#| echo: true

ggplot(dat) +
  aes(x = longitude, y = latitude, colour = richness_raw) +
  geom_point()
```

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| echo: true
suppressMessages(
ggplot(dat, aes(x = latitude, y = richness_raw)) +
  stat_smooth() +
  geom_point()
)
```

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| echo: true
#| output: false

sdat <- st_as_sf(dat, coords = c("longitude", "latitude"),
                 crs = 4326)

crs4326 <- st_crs(4326)
crs4326                 # Look at the whole coordinate reference system
crs4326$Name            # Pull out just the name of the CRS
```

```{r}
#| message: false
#| output: true
#| code-fold: true
#| echo: true

plot(sdat["richness_raw"])
```

```{r}
#| message: false
#| code-fold: true
#| echo: true
#| output: false

# Using tmap

tm1 <- tm_shape(sdat) +
  tm_dots(
    fill = "richness_raw",
    fill.scale = tm_scale_continuous(values = "plasma"),
    fill.legend = tm_legend(title = "Species richness"),
    size = 0.6,
    col = "black",
    lwd = 0.3
  ) +
  tm_title("Species Richness") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE
  )

# Loading shapefiles

aus <- st_read("../data/workshop4/spatial-data/Aussie/Aussie.shp")
aus

shelf <- st_read("../data/workshop4/spatial-data/aus_shelf/aus_shelf.shp")
shelf


```

### Map Design and Aesthetics

```{r}
#| message: false
#| code-fold: true
#| echo: true

# Mapping polygons

tm_shape(shelf) +
  tm_polygons()
```

```{r}
#| message: false
#| output: true
#| code-fold: true
#| echo: true

tm_shape(shelf, bbox = sdat) +
  tm_borders(
    fill = "grey90",
    col = "grey50",
    lwd = 1,
    lty = "dashed"
  ) +
  
  tm_shape(aus) +
  tm_polygons(
    fill = "grey80",
    col = "grey30",
    lwd = 1
  ) +
  
  tm_shape(sdat) +
  tm_dots(
    size = 0.02,
    col = NA
  )
```

### Reflection on Design Choices

I opted to keep the colours relatively plain and greyscale, as I feel this looks moreprofessional and is easier on the eye. The map still conveys the information effectivelywithout unnecessary visual distractions. However, if the map were to include more data,I might consider using additional colours or visual elements to distinguish betweendifferent features and improve readability.
